RED WINE EDA
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(gridExtra)
library(dplyr)
library(RColorBrewer)
library(GGally)
library(lintr)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data

wines <- read.csv("wineQualityReds.csv")
names(wines)
```

This tidy data set contains 1,599 red wines with 11 variables on the 
chemical properties of the wine. At least 3 wine experts rated the 
quality of each wine, providing a rating between 0 (very bad) and 
10 (very excellent).

   Input variables (based on physicochemical tests):
   1 - fixed acidity (tartaric acid - g / dm^3)
   2 - volatile acidity (acetic acid - g / dm^3)
   3 - citric acid (g / dm^3)
   4 - residual sugar (g / dm^3)
   5 - chlorides (sodium chloride - g / dm^3
   6 - free sulfur dioxide (mg / dm^3)
   7 - total sulfur dioxide (mg / dm^3)
   8 - density (g / cm^3)
   9 - pH
   10 - sulphates (potassium sulphate - g / dm3)
   11 - alcohol (% by volume)
   Output variable (based on sensory data): 
   12 - quality (score between 0 and 10)


Which chemical properties influence the quality of red wines?

# Univariate Plots Section

This report explores a dataset containing quality and attributes of 1599 red wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wines)
```

Our dataset consists of eleven variables, with almost 1599 observations.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# function for univariate histogram

univariate_plot <- function(varname, binwidth) {
return(ggplot(aes_string(x = varname), data = wines) + 
         geom_histogram(binwidth = binwidth))
}

```

```{r echo=FALSE, message=FALSE, warning=FALSE}  
univariate_plot ('fixed.acidity', 0.1)
summary(wines$fixed.acidity)
```

Fixed Acidity - nominal distribution with long right tail aka right/positive skewness

```{r echo=FALSE, message=FALSE, warning=FALSE}  

univariate_plot ('volatile.acidity', 0.1)
summary(wines$volatile.acidity)

```

Volatile Acidity - nominal distribution with slight positive skewness and 
possible outlier

```{r echo=FALSE, message=FALSE, warning=FALSE}  
univariate_plot ('citric.acid', .01)
summary(wines$citric.acid)

```

Citric Acid - lots of red wines with 0 citric acid, many with just under 0.5

```{r echo=FALSE, message=FALSE, warning=FALSE}  
univariate_plot ('chlorides', .01)
summary(wines$chlorides)
```

Chlorides - normal distribution but long right tail, right skewed,
possible outliers

```{r echo=FALSE, message=FALSE, warning=FALSE}  
univariate_plot ('free.sulfur.dioxide', 1)
summary(wines$free.sulfur.dioxide)
```

Free Sulfur Dioxide - long right tail, heavy right skewedness,
possible outliers

```{r echo=FALSE, message=FALSE, warning=FALSE}  
univariate_plot ('total.sulfur.dioxide', 1) +
  scale_x_continuous(limits = c(0, 170))
summary(wines$total.sulfur.dioxide)
```

Total Sulfur Dioxide - long right tail, heavy right skewedness,
possible outliers (max value 289)

```{r echo=FALSE, message=FALSE, warning=FALSE}  
univariate_plot ('density', .0001)
summary(wines$density)
```

Density - normal distribution

```{r echo=FALSE, message=FALSE, warning=FALSE}  
univariate_plot ('pH', 0.1)
summary(wines$pH)

```

pH - normal distribution

```{r echo=FALSE, message=FALSE, warning=FALSE}  
univariate_plot ('sulphates', .01) +
  scale_x_continuous(limits = c(0.2, 1.7))
summary(wines$sulphates)
```

Sulphates - normal distribution, slight right skewness, 
possible outliers (max value 2)

```{r echo=FALSE, message=FALSE, warning=FALSE}  
univariate_plot ('alcohol', 0.1)
summary(wines$alcohol)

```

Alcohol - right skewness with a few wines with very low alcohol and 
some with very high alcohol

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality), data = wines) +
  geom_bar()
summary(wines$quality)
```

The distribution of Red Wine Quality appears to be normal. A large majority of Red Wine in the dataset were given 5 or 6 Quality scores. There are very few Red Wines with scores 3 or 8. There are no scores less than 3 or greater than 8.

# Univariate Analysis

### What is the structure of your dataset?
There are 1599 red wines in the dataset with 11 attributes:
   1 - fixed acidity (tartaric acid - g / dm^3)
   2 - volatile acidity (acetic acid - g / dm^3)
   3 - citric acid (g / dm^3)
   4 - residual sugar (g / dm^3)
   5 - chlorides (sodium chloride - g / dm^3
   6 - free sulfur dioxide (mg / dm^3)
   7 - total sulfur dioxide (mg / dm^3)
   8 - density (g / cm^3)
   9 - pH
   10 - sulphates (potassium sulphate - g / dm3)
   11 - alcohol (% by volume)

### What is/are the main feature(s) of interest in your dataset?
Quality is the main feature because we are investigating which chemical properties influence the quality of red wines.  
Quality looks like a normal distribution, lots of 5 and 6 rankings, less 4 and 7, very few 3 and 8

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
- some wines have no citric acid
- most wines between 9-12% alcohol
- wines sulphur dioxide content is long tailed
- free sulfur dioxide appears to correlated to total sulfur dioxide

### Did you create any new variables from existing variables in the dataset?
- Not yet... 

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
- Not yet... 

# Bivariate Plots Section

First let's calculate all correlations instead of one at a time
Using ggcor

```{r echo=TRUE, message=FALSE, warning=FALSE}  
# calculate all correlations instead of one at a time
ggcorr(wines, label=TRUE, hjust = 0.75, size = 3, color = "grey50", label_size = 3, layout.exp = 1, label_round = 2, label_alpha = TRUE)
```

The following variables have a less than 0.20 correlation with quality:
fixed acidity,
residual sugar,
chlorides,
free sulfur dioxide,
total sulfur dioxide, 
density,
pH

Lets plot scatter and boxplots for variables which seem to 
correlate with quality

```{r echo=FALSE, message=FALSE, warning=FALSE}
# function for bivariate scatterplots

bivariate_plot <- function(yvar) {
return(ggplot(aes_string(x = factor(wines$quality), y = yvar), 
              data = wines) + 
         geom_jitter(alpha = 0.1) + 
         geom_boxplot(alpha = 0.1, color = "blue") +
         stat_summary(fun.y = "mean", 
                      geom = "point", 
                      color = "red") +
         labs(x = "Quality"))
}

```


```{r echo=FALSE, message=FALSE, warning=FALSE}
# function for bivariate scatterplots
bivariate_plot("volatile.acidity")
```

volatile acidity appears to negatively correlate with quality

```{r echo=FALSE, message=FALSE, warning=FALSE}

bivariate_plot("citric.acid")
```

citric acid appears to positively correlate with quality
again we see lots of red wine with 0 citric acid
higher quality wines generally do not have 0 citric acid

```{r echo=FALSE, message=FALSE, warning=FALSE}

bivariate_plot("sulphates") + 
  scale_y_continuous(limit = c(0.3,1.1))
```

Sulphates appears to be positively correlated with quality
lots of outliers with very high sulphates

```{r echo=FALSE, message=FALSE, warning=FALSE}

bivariate_plot("alcohol")
```

Alcohol appears to be negatively correlated with quality
Higher Quality scores generally had higher alcohol

total sulphur dioxide and density have correlation factors nearly 0.2
let's graph them just to double check.

```{r echo=FALSE, message=FALSE, warning=FALSE}

bivariate_plot("total.sulfur.dioxide") +
  scale_y_continuous(limits = c(0, 170))
```

not much correlation for total.sulfur.dioxide
quality scores of 5 and 6 had highest mean total.sulfur.dioxide
low quality and high quality red wine had lower mean total.sulfur.dioxide 

```{r echo=FALSE, message=FALSE, warning=FALSE}

bivariate_plot("density")
```

We do observe a slight negative correlation between
density and quality

Of the attributes that were correlated with quality,
Volatile Acidity was also correlated with
Citric acid and Sulphates

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes_string(x = wines$volatile.acidity,
                  y = wines$citric.acid), 
       data = wines) + 
  geom_jitter(alpha = 0.1) + 
  geom_smooth(method = "lm") +
  labs(x = "volatile.acidity",
       y = "citric.acid") +
  scale_x_continuous(limits = c(0.0, 1.1)) +
  scale_y_continuous(limits = c(0.0, 0.8))
```

Strong negative correlation between citric acid and volatile acidity
Red wines that had 0 citric acid had higher volatile acidity
  
```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes_string(x = wines$volatile.acidity,
                  y = wines$sulphates), 
       data = wines) + 
  geom_jitter(alpha = 0.1) + 
  geom_smooth(method = "lm") +
  labs(x = "volatile.acidity",
       y = "sulphates") +
  scale_x_continuous(limits = c(0.0, 1.1)) +
  scale_y_continuous(limits = c(0.3, 1.5))
```

Weak negative correlation between volatile acidity and sulphates

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes_string(x = wines$density,
                  y = wines$alcohol), 
       data = wines) + 
  geom_jitter(alpha = 0.1) + 
  geom_smooth(method = "lm") +
  labs(x = "density",
       y = "alcohol")
```

Very strong negative correlation betwen alcohol and density

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Volatile Acidity and Quality (-0.39)
Citric Acid and Quality (+0.23)
Density and Quality (-0.19)
Sulphates and Quality (+0.25)
Alcohol and Quality (+0.48)

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Volatile Acidity and Citric Acid (-0.55)
Volatile Acidity and Sulphates (-0.26)
Density and Alcohol (-0.50)

Fixed Acidity and Density (+0.67)
Fixed Acidity and pH (-0.68)
Citric Acid and pH (-0.54)
Free Sulphur Dioxide and Total Sulpher Dioxide (0.67)


### What was the strongest relationship you found?

The strongest relationship with Quality were:

alcohol (0.476)
Volatile Acidity (-0.390)
sulphates (0.251)
Citric Acid (0.226)
Density (-0.17)

# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE}
# The very good red wines had low volatile acidity and high citric acid 
ggplot(aes(x = density, y = alcohol, 
           color=factor(quality)), data = wines) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
  scale_x_continuous(limits = c(0.990, 1.005)) +
  scale_y_continuous(limits = c(9, 14)) +
  ggtitle('Quality by Density and Alcohol') +
  theme_dark()
```

Many of the very high quality red wines are low in density
and high in alcohol

```{r echo=FALSE, message=FALSE, warning=FALSE}
# The very good red wines had low volatile acidity and high citric acid 
ggplot(aes(x = volatile.acidity, y = citric.acid, 
           color=factor(quality)), data = wines) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
  scale_x_continuous(limits = c(0.2, 1.1)) +
  scale_y_continuous(limits = c(0, 0.8)) +
  ggtitle('Quality by Volatile Acidity and Citric Acid') +
  theme_dark()
```

Higher quality wines appear to be high in citric acid and
low in volatile acidity

```{r echo=FALSE, message=FALSE, warning=FALSE}
# The very good red wines had low volatile acidity and high sulphates 
ggplot(aes(x = volatile.acidity, y = sulphates, 
           color=factor(quality)), data = wines) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
  scale_x_continuous(limits = c(0.2, 1.0)) +
  scale_y_continuous(limits = c(0.3, 1.1)) +
  ggtitle('Quality by Volatile Acidity and Sulphates') +
  theme_dark()
```

Higher quality wines appear to be high in sulphates and
low in volatile acidity

```{r echo=FALSE, message=FALSE, warning=FALSE}
# The higher quality red wines generally had high alcohol and low volatile acid 
ggplot(aes(x = alcohol, y = volatile.acidity, 
           color=factor(quality)), data = wines) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
  scale_x_continuous(limits = c(9, 14)) +
  scale_y_continuous(limits = c(0.3, 1.1)) +

  ggtitle('Quality by Alcohol and Volatile Acid') +
  theme_dark()
```

Higher quality wines appear to be high in alcohol and
low in volatile acidity

```{r echo=FALSE, message=FALSE, warning=FALSE}
# The higher quality red wines generally had high alcohol and high sulphates

ggplot(aes(x = alcohol, y = sulphates, 
           color=factor(quality)), data = wines) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
  scale_x_continuous(limits = c(9, 14)) +
  scale_y_continuous(limits = c(0.3, 1.1)) +
  ggtitle('Quality by Alcohol and Sulphates') +
  theme_dark()
```

Higher quality wines appear to be high in alcohol and
high in sulphates

```{r echo=FALSE, message=FALSE, warning=FALSE}
# The higher quality red wines generally had high alcohol and high citric acid

ggplot(aes(x = alcohol, y = citric.acid, 
           color=factor(quality)), data = wines) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = "seq",
    guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
  scale_x_continuous(limits = c(9, 14)) +
  scale_y_continuous(limits = c(0.0, .8)) +
  ggtitle('Quality by Alcohol and Citric Acid') +
  theme_dark()
```

Higher quality wines appear to be high in alcohol and
high in citric acid

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

The combination of high Sulphates and high alcohol seems to result in the highest Quality scores.

Both citric acid and volatile acidity had correlations with quality of red wine. 
Volatile acidity negatively correlated with quality. 
Citric acid positively correlated with quality. 
Citric acid is also negatively correlated with volatile acidity and 
positively correlated with fixed acidity.



### Were there any interesting or surprising interactions between features?

Residual sugar was not correlated with Quality scores

### OPTIONAL: Did you create any models with your dataset? Discuss the \ 
strengths and limitations of your model.

------

# Final Plots and Summary

### Plot One
```{r echo=TRUE, message=FALSE, warning=FALSE, Plot_One}
ggplot(aes(x = quality), data = wines) +
  geom_bar() +
  xlab("Quality") +
  ylab("Number of Red Wines") +
  ggtitle("Red Wine Quality counts")
```

### Description One
The distribution of Red Wine Quality appears to be normal. A large majority of Red Wine in the dataset were given 5 or 6 Quality scores. There are very few Red Wines with scores 3 or 8. There are no scores less than 3 or greater than 8.

### Plot Two
```{r echo=TRUE, message=FALSE, warning=FALSE, Plot_Two}
bivariate_plot("alcohol") +
  xlab("Quality") +
  ylab("Alcohol Percentage (%)") +
  ggtitle('Alcohol Percentage (%) by Quality')
```

### Description Two
Red Wines with Quality of 3, 4, and 5 had low Alcohol Percentages, median around 10%. Increasing Quality Red Wines had increasing median Alcohol Percentages.

### Plot Three
```{r echo=TRUE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(aes(x = alcohol, y = sulphates, 
           color=factor(quality)), data = wines) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
  scale_x_continuous(limits = c(9, 14)) +
  scale_y_continuous(limits = c(0.3, 1.1)) +
  theme_dark() +
  ggtitle('Quality by Alcohol and Sulphates') +
  xlab("Alcohol Percentage (%)") +
  ylab("potassium sulphate (g/dm3)")
```

### Description Three
Red Wines with Quality score of 3 typically had low Sulphates and low Alcohol Percentages. Higher levels of Sulphates and Alcohol Percentages resulted in higher scoring Quality in red wines.


------

# Reflection

The Red Wines data set contains information on almost 1,599 red wines. I started by understanding the individual variabltes in the data set using plots. Using many R libraries, I was able to determine which variables had a greatest impact on Quality and which variables were correlated.

I learned that citric acid correlated with quality of red wine. Citric acid is also negatively correlated with volatile acidity and positively correlated with fixed acidity. Any models would need to take this into account.

I struggled with figuring out which type of graph was best suited for this investigation, which resulted in creating more graphs than necessary, re-creating graphs and deleting graphes. I suspect with more experience I could spend more time planning which graphs I would need to draw conclusions instead of simply starting with graphs.

Future work could include further analysis and developing a model to predict Quality scores.


# References

https://www.statmethods.net/management/subset.html
https://stats.stackexchange.com/questions/177129/ggplot-and-loops

https://ourcodingclub.github.io/2017/02/08/funandloops.html
http://rprogramming.net/rename-columns-in-r/

https://stackoverflow.com/questions/10085806/extracting-specific-columns-from-a-data-frame

for corr
http://www.sthda.com/english/wiki/correlation-matrix-a-quick-start-guide-to-analyze-format-and-visualize-a-correlation-matrix-using-r-software 
https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html
https://briatte.github.io/ggcorr/ 
https://bibinmjose.github.io/RedWineDataAnalysis/#correlation_matrix 

for colorbrewer2
https://stackoverflow.com/questions/31297196/continuous-value-supplied-to-discrete-scale
http://ggplot.yhathq.com/docs/scale_color_brewer.html
http://colorbrewer2.org/#type=diverging&scheme=Spectral&n=3 

comments on submission 1
https://review.udacity.com/#!/reviews/1398788

for lintr
https://github.com/jimhester/lintr/blob/master/README.md

